### 2016-08-10 doc/publications/plos16/plos16.tex
Some changes

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{Cursed Forest - A random forest implementation for `big' and `wide' data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Aidan O'Brien\textsuperscript{1},
Piotr Szul\textsuperscript{2},
Stephanie Li\textsuperscript{3},
James Doecke\textsuperscript{3},
Nick Ellis\textsuperscript{4},
Robert Dunne\textsuperscript{5}, and
Denis C. Bauer\textsuperscript{1,*}
%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\bf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
\\
\bf{2} Data61, CSIRO, Brisbane, QLD, Australia
\\
\bf{3} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
\\
\bf{4} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
\\
\bf{5} Data61, CSIRO, Sydney, NSW, Australia
\\
\bigskip

@@

\section*{Notes}
\begin{itemize}
\item what are you guys editing this with?  what about some carriage
  returns or line feeds or something!
\item Are we going to use the TCGA data in this paper?
\item are we going to run sparkML or Cursed forest on the extended
  simulation data \cite[]{Genuer.et.al.2010}? Has Aidan done this already?
\item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
  \ref{table:smaller_data} -- On it, Aidan
\item check Piotr's simulation
\item run an argument as follows:
  \begin{itemize}
  \item variable  selection may succeed where accurate prediction does not
  \item consider  \cite{Segal.2004} examples
  \item  perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form we can recover
    the reltionship for very large $p$p
  \end{itemize}
\item \cite{Tuv.et.al.2009} feature selection via adding permuted variables. Do we want to consider this?
\item check out varSelRF
I think the structure of the first section should be
\begin{itemize}
\item  convergence results from Biau 2012
\item simulations reproduced from Genuer et al 2010
\item we extend the simulations to many more variables
\item discussion of the effect of the ntree and mtry parameters for variable selection with many noise variables
\item  in the case of some functions (step, ramp) we can recover the significant variables in the presence of a very large number of noise variables
\item in the case of genomic data, our functions of interest are more likely to be stop or ramp functions rather than harmonic functions. So we have some hope.
\end{itemize}

\item check out 
  \begin{itemize}
  \item D\'iaz-Uriarte, R., \& Alvarez de Andr\'es, S. (2006). Gene selection and classification of microarray data using random
    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
  \item Goldstein et al. (2011)
  \item Chen (2012)
  \item VSURF and  varSelRF, 
  \item ranger package
  \item  r2VIM: A new variable selection method for random forests in
    genome-wide association studies
  \item CloudForest: A Scalable and Efficient Random Forest Implementation for Biological Data
  \end{itemize}
\end{itemize}

@@

The digital revolution is seeing a dramatic increase in data collected about every aspect of 
life~\cite{Loebbecke2015}.  These dataset not only grow vertically by capturing more and more events but also
horizontally by capturing more information about these events.  The challenge of big and `wide' data is especially
pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3
billion basepairs of the human genome.  While a large fraction of this data is identical between individuals, \marginpar{base-pairs}
identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].

For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  The richness of
this data can be utilised in other datasets as well by filling in or ``imputing" the information at genomic positions
previously unobserved due to the less detailed resolution of older array technology [citation about imputation].  This
means that the large number of samples for which genome-wide association (GWA) data is available~\cite{Welter2013} can
be extended horizontally by imputing the previously unobserved genotypes.  This provides the potential to generate
datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating
modern compute paradigms to deal with these challenges.

@@

% Developed b Apache, Hadoop is one platform for dealing with large data sets is Platforms Fortunately computational
% tools such as Apache Spark are available as a framework for iteratively dealing with big data.
Distributing the compute tasks is one strategy to overcome this challenge. Apache Spark, based on Scala, allows for  
\marginpar{is it based on Scala or does it have a Scala API?}
distributed applications to scale to thousands of nodes.  A Spark application runs on a `driver' node, and
this node divides and distributes work out to the many `worker' nodes.  Spark also includes the machine learning (ML)
library, Spark ML, providing popular machine learning libraries.  Spark ML includes supervised approaches such as
decision trees and the ensemble method, random forests, as well as unsupervised approaches such as k-means clustering.

@@

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
supervised learning, i.e. given a phenotype of clinical label determine the most associated features.  One of the
simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
linear combination of the feature variables.  Processing the `wide' genomic data poses difficulties for machine learning
applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
This is known as the `curse of dimensionality'~\cite{Bauer2014}.  
\marginpar{is this the reference we want for the ``curse of dimensionality''  Bellman (1961) is a more ususl reference}

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are scale invariant;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to useless predictors;
  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model.

\subsection*{RF}

@@

The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
the $i=1,..,Q$ classes
\begin{equation*}
{\bar {h}}=  \arg \max_i \left(\sum_b I(h(x;\theta_b)=i)\right).
\end{equation*}

Random forests are a variance reduction technique. They take a tree models, which have a low bias but a high variance, and
by combining them reduce the variance. The prediction error is the sum of the variance and the bias squared.

@@

%http://stats.stackexchange.com/questions/77018/random-forest-is-it-a-boosting-algorithm
 %http://stats.stackexchange.com/questions/173390/gradient-boosting-tree-vs-random-forest
%overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
%between features, which is of importance for modelling complex polygenic diseases.

@@

The benefits of random forests models are
  \begin{itemize}
  \item they give an out-of-bag estimate of model accurace;
  \item they provide a measure of variable importance;
  \item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
    variables. This would be difficult as the trees as split on random subsets of variables).
  \end{itemize}

Since the original implmentation of random forests there have been a number of developments leading to a better understanding of
the algorithm. 
\begin{itemize}
\item random forests are resistent against overfitting. However it is not true that they will not overfit; see \cite{Segal.2004}
  where an example of RF overfitting is  given, caused by the fact that successive trees were correlated;
\item \cite{Strobl.et.al.2007} demonstrates that the RF algorithm is subject to a bias in variable selection via two mechanisms:
  \begin{enumerate}
  \item variable scale (for $X_I$ continuous) and number of levels (for $X_i$ discrete) inroduces a bias. A uninformative  variable with a large
    number of levels may be selected over a more informative  variable with fewer levels;
  \item bagging introduces a bias, but sampleing about $(1- 1/e) \approx 0.632$  of the data without 
  \end{enumerate}
\end{itemize}

\cite{Strobl.et.al.2007} introduce changes to the RF algorithm to fix 1) and 2). 
However the level effect will be apparent in the CF algorithm as well. \marginpar{we can fix the boosting effect}
However, there are many examples, (particularly in genomics) whre we have data sets that are
\begin{itemize}
\item very wide
\item all variables have the same number of levels (measures at different points along the genome)
\item the hypothesied functional relationship is of a form amenable to a DT.
\end{itemize}

However, we observe that Spark's standard RF implementation is not able to handle the extremely `wide' genomic data as
it was developed for large number of samples with only modest dimensionality.  Although Spark ML can build a RF model on
a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
scaling to larger datasets.

% The reason for the failure can be attributed to the "curse of dimensionality". Where Spark ML is designed to process a
% huge number of samples with a modest dimensionality, the 1000 Genomes data instead contains a huge number of
% dimensions (features). As a result, the default random forests implementation is unable to partition the data into
% small-enough tasks for nodes on the computer cluster to handle.

We therefore developed CursedForest, a Spark-based random forest algorithm able to process big and `wide' data. Unlike
the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
excessive resource requirements on the driver or worker nodes.
%Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
In the first section we

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis"

\subsubsection*{variable selection}
\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using the
importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic data.
\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of RF to handle GWAS data. Waht do we nned to do in this area}

\section*{Methods}

\subsection*{VariantSpark.}
Traditionally, VariantSpark made use of Spark's machine learning algrorithms `Spark ML'. These algorithms require data in a `DataFrame',
similar to DataFrames in R.

@@

Cursed Forest uses:
\begin{itemize}
\item an ``information gained'' criteria for splitting.  $f_q$ be the fraction of items labeled with value $q$ where $q= 1,
  \ldots, Q$ in a set. The  entropy of the set is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i$
The expected information gain is the change in information entropy $H$ from a prior state $T$ to a state that takes some
information $a$ as given: 
\[ IG(T,a) = H(T) - H(T|a) \]
In this case, the  the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
\item variable importance is the sum of IG over all notes and all trees for that variable
\item the model of parallelism is different to the ``foreach'' method -- we need to describe it
\end{itemize}

@@

\marginpar{These are regression examples, we nned some classification examples}

@@

Traditionally, VariantSpark used the random forest algorithm from SparkML. While SparkML algorithms demonstrate
scalability when dealing with a large number of samples, this scalability quickly breaks down as we include more features.

One of the restrictions preventing SparkML algorithms from scaling to high-dimensionality data is the fact that the feature-set
for each individual is stored as a `Vector'. In Spark, Vectors are unable to be split and to process multiple vectors together,
the entirety of each vector must be loaded into memory on the same node. Therefore, as dimensionality increases, the
size of these Vectors must also increase, and subsequently this results in higher memory requirents.

On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed" data. It avoids the relation between
memory and dimensionality, by not using Vectors. Instead, CursedForest stores data in the Parquet columnar format.

\begin{table}[!ht]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l+l|l|l|l|l|l|l|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
\bf{Genome \%}  & \bf{method} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest &  &  &  \\
& RF (Spark ML) &  &  &  \\
& RF (R) & 0.456 & 2.89 & 18GB \\ \hline

\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364} & CursedForest & XX & 1hr 58min 23sec & 4GB \\
& RF (Spark ML) & 0.94 & 8hr 29min 07sec & 8GB \\
& LR &  & 10hr 30min & 8GB\\ \hline

\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051}  & CursedForest & &  &\\
& RF(Spark ML) & &  &\\
& LR & &  &\\ \hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & 0.96 & 7hr 14min 40sec & 8GB \\ 
& RF (Spark ML) & - & - & - \\
& LR & - & - & - \\ 
& kmeans (Spark ML) & 0.82 & 30hr 44min 00sec & 24GB \\ \hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project. Size of the data is individuals x variants

@@

\subsection*{Parameter settings}
We consider the parameter setting for the RF algorithm. We use the R terms from the package RandomForests
\cite{Liaw.and.Weiner.2002} which incoporates the original Fortran code by Brieman and Cutler. We incprporate the 
advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience.

\begin{itemize}
\item \texttt{ntree} the number of trees.  The number of trees necessary for good performance grows with the number of predictors.
  \marginpar{can we do proxmiity?}  \cite{Liaw.and.Weiner.2002} suggest that a high \texttt{ntree} is necessary to get stable
  estimates of variable importance and proximity; however, even though the variable importance measures may vary from run to run,
  the ranking of the importances is quite stable.  In addition, we note that it is possible for a RF model to have a very poor
  fit, and be useless for prediction, and still have a useful ranking of variable importance.
\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decison
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \texttt{mtry} may give
  better performance.
\item  For classification problems where the class frequencies are extremely unbalanced it may be necessary to change the prediction
  rule \marginpar{can we do this}
\item the size and complexity of the individual trees is controlled in RandomForests by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark.ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest. The SparkML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
  example\footnote{The documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html}.}.

  In the ATCG example the RandomForests code produced trees with depths of 30+. We note that Random forests are a way of averaging
  multiple deep decision trees, trained on different parts of the same training set, with the goal of reducing the
  variance. \url{https://en.wikipedia.org/wiki/Random_forest} and refs therein.  
\marginpar{are we including this  example?  Supplementary info?}
\end{itemize}

@@

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Sensitivity analysis on parameter choice.}}
\begin{tabular}{|l|l|l|l|l|l|l|l|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
\bf{Mtry}  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
\hline
&&&&&\\ \hline
\end{tabular}
\begin{flushleft} 
  Table notes Phasellus venenatis, tortor nec vestibulum mattis, massa tortor interdum felis, nec pellentesque metus
  tortor nec nisl. Ut ornare mauris tellus, vel dapibus arcu suscipit sed.
\end{flushleft}
\label{table2}
%\end{adjustwidth}
\end{table}

@@

\subsection*{Conclusion}

@@

% Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
\subsection*{S1 Video}
\label{S1_Video}
{\bf Bold the first sentence.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\subsection*{S1 Text}
\label{S1_Text}
{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

\subsection*{S1 Table}
\label{S1_Table}
{\bf Lorem Ipsum.} Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

@@

  \item D\iaz-Uriarte, R., \& Alvarez de Andr\es, S. (2006). Gene selection and classification of microarray data using random
    forest. BMC Bioinformatics, 7, 3. http://doi.org/10.1186/1471-2105-7-3
  \item Goldstein, B. A., Polley, E. C., & Briggs, F. B. S. (2011). Random Forests for Genetic Association Studies. Statistical
    Applications in Genetics and Molecular Biology, 10(1), 32. http://doi.org/10.2202/1544-6115.1691

@article{Chen2012323,
title = "Random forests for genomic data analysis ",
journal = "Genomics ",
volume = "99",
number = "6",
pages = "323 - 329",
year = "2012",
note = "",
issn = "0888-7543",
doi = "http://dx.doi.org/10.1016/j.ygeno.2012.04.003",
url = "http://www.sciencedirect.com/science/article/pii/S0888754312000626",
author = "Xi Chen and Hemant Ishwaran",
keywords = "Random forests",
keywords = "Random survival forests",
keywords = "Classification",
keywords = "Prediction",
keywords = "Variable selection",
keywords = "Genomic data analysis ",
abstract = "Random forests (RF) is a popular tree-based ensemble machine learning tool that is highly data adaptive, applies to “large p, small n” problems, and is able to account for correlation as well as interactions among features. This makes \{RF\} particularly appealing for high-dimensional genomic data analysis. In this article, we systematically review the applications and recent progresses of \{RF\} for genomic data, including prediction and classification, variable selection, pathway analysis, genetic association and epistasis detection, and unsupervised learning. "
}

>>>>

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{Cursed Forest - A random forest implementation for ``big'' and ``wide'' data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Aidan O'Brien\textsuperscript{1},
Piotr Szul\textsuperscript{2},
Stephanie Li\textsuperscript{3},
James Doecke\textsuperscript{3},
Nick Ellis\textsuperscript{4},
Robert Dunne\textsuperscript{5}, and
Denis C. Bauer\textsuperscript{1,*}
%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\bf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
\\
\bf{2} Data61, CSIRO, Brisbane, QLD, Australia
\\
\bf{3} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
\\
\bf{4} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
\\
\bf{5} Data61, CSIRO, Sydney, NSW, Australia
\\
\bigskip

@@

\section*{Notes}
\begin{itemize}
\item what are you guys editing this with?  what about some carriage
  returns or line feeds or something!
\item Are we going to use the TCGA data in this paper?
\item are we going to run spark ML or Cursed forest on the extended
  simulation data \cite[]{Genuer.et.al.2010}? Has Aidan done this already?
\item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
  \ref{table:smaller_data} -- On it, Aidan
\item check Piotr's simulation
\item run an argument as follows:
  \begin{itemize}
  \item variable  selection may succeed where accurate prediction does not
  \item consider  \cite{Segal.2004} examples
  \item  perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form we can recover
    the reltionship for very large $p$p
  \end{itemize}
\item \cite{Tuv.et.al.2009} feature selection via adding permuted variables. Do we want to consider this?
\item chek out varSelRF
\end{itemize}

@@

The digital revolution or ``datafication'' sees a dramatic increase in data collected about every aspect of   \marginpar{is seeing, will see}
life~\cite{Loebbecke2015}.  These dataset not only grow vertically by capturing more and more events but also
horizontally by capturing more information about these events.  The challenge of big and ``wide'' data is especially
pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3
billion basepairs of the human genome.  While a large fraction of this data is identical between individuals, \marginpar{base-pairs}
identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].

For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012},
denoting the differences between individuals.  This dataset consumes close to a terabyte of disk space.  The richness of
this data can be utilised in other datasets as well by filling in or ``imputing'' the information at genomic positions
previously unobserved due to the less detailed resolution of older array technology [citation about imputation].  This
means that the large number of samples for which genome-wide association (GWA) data is available~\cite{Welter2013} can
be extended horizontally by imputing the previously unobserved genotypes.  This provides the potential to generate
datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating
modern compute paradigms to deal with these challenges.

@@

% Developed b Apache, Hadoop is one platform for dealing with large data sets is Platforms Fortunately computational
% tools such as Apache Spark are available as a framework for iteratively dealing with big data.
Distributing the compute tasks is one strategy to overcome this challenge. Apache Spark, based on Scala, allows for  
\marginpar{is it based on scale or does it have a scale API?}
distributed applications to scale to thousands of nodes.  A Spark application runs on a ``driver'' node, and this node
divides and distributes work out to the many ``worker'' nodes.  Spark also includes the machine learning (ML) library,
Spark ML, providing support for a selection of machine learning algorithms.  Spark ML includes supervised approaches such
as decision trees and the ensemble method, random forests, as well as unsupervised approaches such as k-means clustering.

@@

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is
supervised learning, i.e. given a phenotype of clinical label determine the most associated features.  One of the
simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a
linear combination of the feature variables.  Processing the ``wide'' genomic data poses difficulties for machine learning
applications as having more features $p$ than samples $n$ causes the model to overfit the data and generalize poorly.
This is known as the ``curse of dimensionality''~\cite{Bauer2014}.  
\marginpar{is this the reference we want for the ``curse of dimensionality''}

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are scale invariant;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to useless predictors;
  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model. 
\marginpar{a potential problem  with tree models is the tendencey to split on categorical variables with a high number of
  levels, even when better splits are available}
\subsection*{Random Forest}

@@

The results are combined in an appropriate way, i.e. for a regression problem,
\begin{equation*}
{\bar {h}}={\frac {1}{B}}\sum _{b=1}^{B} h(x;\theta_b).
\end{equation*}
\marginpar{make it a classification example}
%where $x$ is the covariate vector of length $p$
%with associated random vector $X$ and the $\theta_b$ are i.i.d. random vectors.

@@

\subsection*{Benefits of RF}

\marginpar{weak learners only in a misuse of the term. ``weak learners'' was applied to decision stumps}
%http://stats.stackexchange.com/questions/77018/random-forest-is-it-a-boosting-algorithm
%http://stats.stackexchange.com/questions/173390/gradient-boosting-tree-vs-random-forest
%overcomes this issue and is hence ideally suited to this application.  Furthermore, RF are able to capture interactions
%between features, which is of importance for modelling complex polygenic diseases.

@@

Random forests are a variance reduction technique. They take a tree models which have a low bias but a high variance and
by combining them reduce the variance. The prediction error is the sum of the variance and the bias squared.

The benefits of random forests models are:
  \begin{itemize}
  \item they give an out-of-bag estimate of model accurace;
  \item they provide a measure of variable importance;
  \item they inherit many of the qualities of tree models.
\marginpar{but not the handling of missing velues via surrogate variables. This would be difficult as the trees are split
  on random subsets of variables}
  \end{itemize}

However, we observe that Spark ML's standard RF implementation is unable to handle the extremely ``wide'' genomic data as
it was developed for large number of samples with only modest dimensionality.  Although Spark ML can build a RF model on
a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
Spark ML uses, which we elaborate on in the discussion.

% The reason for the failure can be attributed to the ``curse of dimensionality''. Where Spark ML is designed to process a
% huge number of samples with a modest dimensionality, the 1000 Genomes data instead contains a huge number of
% dimensions (features). As a result, the default random forests implementation is unable to partition the data into
% small-enough tasks for nodes on the computer cluster to handle.

We therefore developed CursedForest, a Spark-based random forest algorithm able to process big and ``wide'' data. Unlike
the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
excessive resource requirements on the driver or worker nodes. It also greatly reduces memory requirements on worker nodes
by partitioning the job differently than Spark ML.
%Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
In the first section we explore some of the properties of random forests, demonstrating that any potential flaws in CursedForest
are an inherent property of Random Forests. Here we provide demonstrations using synthetic data.
We then demonstrate the scalability of CursedForest in respect to the dimensionality of the data, building an RF model on whole-genome
data from the 1000 Genomes Project. Here we report on the accuracy (in terms of the ARI) and the error.
Finally, given the role different parameter values can play in model construction, we explore the effect tuning these parameters
can have on the prediction accuracy of a model.

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Methods}

\subsection*{VariantSpark.}
Traditionally, VariantSpark made use of Spark's machine learning algrorithms, ``Spark ML''. These algorithms require data in a ``DataFrame'',
similar to DataFrames in R.

@@

@@

\subsection*{Random Forests converges on wide data}

As discussed in the Introduction,
\marginpar{I think this has been removed from the introduction}
random forests are resistent against overfitting.
However it is not true that they will not overfit; see \cite{Segal.2004} where examples of RF overfitting are given 
\marginpar{discuss these examples}

\cite{Strobl.et.al.2007} demosnstrate that a RF algorithm is subject to a bias in variable selection via two mechanisms:
\begin{itemize}
\item variable scale (for $X_I$ continuous) and number of levels (for $X_i$ discrete) inroduces a bias
\item boosting introduces a bias.
\end{itemize}
These effects will be apparent in the CF algorithm as well. \marginpar{we can fix the boosting effect}
However, there are many examples, (particularly in genomics) whre we have data sets that are
\begin{itemize}
\item very wide
\item all variables have the same number of levels (measures at different points along the genome)
\item the hypothesied functional relationship is of a form amenable to a DT.
\end{itemize}

@@

Traditionally, VariantSpark used the random forest algorithm from Spark ML. While Spark ML algorithms demonstrate
scalability when dealing with a large number of samples, this scalability quickly breaks down as we include more features.

One of the restrictions preventing Spark ML algorithms from scaling to high-dimensionality data is that the feature-set
for each sample is stored as a vector data type. Spark vectors are stored in resilient distributed dataset (RDDs), and while RDDs
can be split, vectors cannot. So as we add more samples (in our case individuals), Spark ML can simply split the RDD into more
partitions and resource requirements will remain relatively stable. However, as the number of dimensions (variants) increases,
the size of the vectors increases. This results in higher memory-requirements as increasingly large vectors must be loaded
into memory.

On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed'' data. It avoids the relation between
memory and dimensionality, by not using Vectors. Instead, CursedForest stores data in the Parquet columnar format.
To build a tree, CursedForest will distribute tasks across compute nodes to calculate the information gain for every feature
in the feature-set. The result is then reduced to return the feature which gives the greatest information gain.
This process is then repeated until CursedForest has calculated every node for said tree.

\begin{table}[!ht]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l+l|l|l|l|l|l|l|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
\bf{Genome \%}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
& RF - Spark ML &  &  &  \\
& RF - R & 0.456 & 2.89 & 18GB \\ \hline

\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364} & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
& RF - Spark ML & 0.94 & 8hr 29min & 8GB \\
& LR - Spark ML&  & 2hr 6min & 8GB\\ \hline

\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051}  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
& RF - Spark ML & &  &\\
& LR - Spark ML & & 14hr 48min & 8GB \\ \hline

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
& RF - Spark ML & - & - & - \\
& LR - Spark ML & - & - & - \\ 
& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ \hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project. Size of the data is listed as individuals x variants.

@@

\subsection*{Parameter settings}
\marginpar{ was``Random forest is sensitive to tree depth''}
In this setting, variable selection in very high dimensions, we nned to consider the parameter setting for the RF
algorithm.

I will use the R terms from the package RandomForests whcih incoporates the original Fortran code by Brieman and
Cutler. We consider the parameters:
\begin{itemize}
\item the size and complexity of the individual trees is controlled in RandomForests by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark.ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest. The Spark ML documentation somewhat unhelpfully sets \texttt{maxDepth=4} in their classification
  example\footnote{the documentation is available at \url{http://spark.apache.org/docs/latest/mllib-ensembles.html} }.
  In the ATCG example the RF code produced trees with depths of 30+. We note that Random forests are a way of averaging
  multiple deep decision trees, trained on different parts of the same training set, with the goal of reducing the
  variance. \url{https://en.wikipedia.org/wiki/Random_forest} and refs therein.  
\marginpar{are we including this  example?  Supplementary info?}
\item \texttt{ntree} the number of trees 
\item \texttt{mtry} --  the number of variables considered at each split (if \texttt{mtry}=$p$, we have a boosted decison
  tree model \marginpar{is this right?}
\end{itemize}

@@

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Sensitivity analysis on parameter choice.}}
\begin{tabular}{|l|l|l|l|l|l|l|l|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
\bf{Mtry}  & \bf{Mtree} & \bf{depth} & \bf{Accuracy} & \bf{Runtime} & \bf{Memory} \\
\hline
&&&&&\\ \hline
\end{tabular}
\begin{flushleft} 
  Table notes here.
\end{flushleft}
\label{table2}
%\end{adjustwidth}
\end{table}

@@

\subsection*{Conclusion}
We find

@@

% Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
%\subsection*{S1 Video}
%\label{S1_Video}
%{\bf Bold the first sentence.}  Maecenas convallis mauris sit amet sem ultrices gravida. Etiam eget sapien nibh. Sed ac ipsum eget %enim egestas ullamcorper nec euismod ligula. Curabitur fringilla pulvinar lectus consectetur pellentesque.

@@
